### 2016-06-23 reasoning.tex
Secs. 4.1 and 4.2 done.

We now describe a proof system that lets us prove that a \txnimp
program satisfies its high-level invariants in all small-step
executions that satisfy a trace invariant\footnote{It is important to
mind the distinction between high-level invariants (\eg~\C{X$\ge$0})
and trace invariants (\eg~\C{RC}). The former constitute \emph{proof}
\emph{obligations} for the programmer, whereas the latter capture
\emph{assumptions} about the machine. } ($\I$).  Since the trace
invariant is a conjunction of store consistency ($\I_s$) and
transaction isolation ($\I_c$) constraints, the proof demonstrates
that program preserves its invariants when executed on a given store
under a given selection of isolation levels for its transactions.

Our proof system generalizes rely-guarantee reasoning to a setting
that includes weakly consistent stores and weakly isolated
transactions.

@@

>>>>

We now describe a proof system that lets us demonstrate that a \txnimp
program satisfies its high-level invariants in all small-step
executions that satisfy a chosen trace invariant\footnote{It is
important to mind the distinction between high-level invariants
(\eg~\C{X$\ge$0}) and trace invariants (\eg~\C{RC}). The former
constitute \emph{proof} \emph{obligations} for the programmer, whereas
the latter capture \emph{assumptions} about the machine. } ($\I$).
Since the trace invariant is a conjunction of store consistency
($\I_s$) and transaction isolation ($\I_c$) constraints, the
demonstration is a proof that the program preserves its invariants
when executed on a chosen store under a selection of isolation levels
for its transactions.

As mentioned earlier, our proof system is a rely-guarantee framework
that admits compositional reasoning by abstracting away the
environmental interference as a rely relation. A conspicuous
difference between a standard development of rely-guarantee and ours
is that, while the former reasons in terms of program states (variable
to value bindings), we reason in terms of executions as captured by
their traces ($\E$). In particular, our rely ($R$) and guarantee ($G$)
relations relate executions (i.e., $R,\,Q\,\subseteq\,\E\times\E$),
and pre ($P$) and post ($Q$) conditions are assertions over executions
(i.e., $P,\,Q\, : \E \rightarrow \Prop$). Our development nonetheless
facilitates state-based reasoning via the interpretation function
($\interp{\cdot}$) introduced in \S\ref{sec:syntax}. For example, if a
\emph{bi-state} rely relation relates every pair of states $\sigma$
and $\sigma'$ such that $\sigma'(X) \ge \sigma(X) \ge 0$, the
corresponding \emph{bi-execution} rely relation relates every pair of
executions $\E$ and $\E'$ such that $\interp{\E}(X) \ge
\interp{\E'}(X) \ge 0$.  Assertions on state are also written
similarly. For instance, consider the postcondition
$\C{C\,=\,k-a1-a2}$ of the program in Fig.~\ref{fig:motiv-eg-1}. The
corresponding assertion on the post-state
($\lambda\sigma.~\sigma(\C{C}) \,=\, \C{k-a1-a2}$) asserts that in all
states resulting from executing the program, the value of \C{C} is
\C{k-a1-a2}. The equivalent execution-based assertion
($\lambda\E.~\interp{\E}(\C{C})\,=\,\C{k-a1-a2}$) asserts that in all
executions of the program, the value of \C{C} is \C{k-a1-a2}.
However, having access to an execution facilitates assertions that go
beyond the state. An example is the invariant at a program point
inside \C{Wd1} described abstractly in \S\ref{sec:motivation}, and
reified as an execution-based assertion below:
\begin{smathpar}
\begin{array}{l}
  \lambda\E.~ \neg(\underE{\committed(\C{Wd1}}) \conj 
      \underE{\C{Wd1} \wrstoar \C{C}}\\
  \hspace*{0.2in}\conj(\neg(\underE{\committed(\C{Wd2})}) \Rightarrow 
      \C{\interp{\E}(C) = k-a1}) \\
   \hspace*{0.2in}\conj (\underE{\committed(\C{Wd2})} \Rightarrow 
      \C{\interp{\E}(C) = k-a1-a2})
\end{array}
\end{smathpar}
% Where, $\committed$ is defined as:
% \begin{smathpar}
% \begin{array}{lcl}
% \underE{\committed(T_i)} & \Leftrightarrow &
%   \exists\eta.~\underE{\eta\in T_i} \conj \kind(\eta) = \C{COMMIT}\\
% \end{array}
% \end{smathpar}
One of the conjuncts asserts that \C{Wd1} has written to \C{C} - a
fact which cannot be deduced solely from the value of \C{C} (esp. in
presence of interference), but can be expressed as a proposition over
the execution trace.

\subsection{The Rely-Guarantee Judgment}

@@

The standard rely guarantee judgment is the quintuple $\rg{P,R}{c}{G,Q}$,
which intuitively means the following: if a command $c$ is executed in
a state that satisfies its precondition $P$, provided that every
interference step during the execution is contained inside the rely
relation $R$, then the effect that each step of executing $c$ has on
the state is captured by $G$, and the final state of execution
satisfies the postcondition $Q$. The terms ``execution'' and ``step''
are defined with respect to an operational semantics, which in our
case, is parameterized on the trace invariant $\I$. Consequently, our
rely-guarantee judgment for commands is the sextuple\footnote{$\I$ is
written in the same group as $P$ to suggest that it is an assumption;
not an obligation.} $\rg{\I,P,R}{c}{G,Q}$, whose semantics differs
from the quintuple in that it requires every step of the execution to
preserve the trace invariant $\I$ for the postcondition to be valid.
As usual, we write $\txnbox{t}_i$ to denote a term $t$ inside a
transaction $T_i$.  \txnimp expressions are side-effecting (they
generate \C{RD} effects), and admit interference during their
evaluation. Therefore, like the commands, expressions need to be
reasoned about explicitly. But, unlike commands, expressions evaluate
to values, and reasoning framework should also admit assertions on such
values. We therefore define a separate judgment for expressions - the
septuple $\rg{\I,P,R}{e}{G,C,Q}$. The new entrant $C : \mathbb{N}
\rightarrow \Prop$ is an assertion on the return value of $e$.

The rules that define rely-guarantee judgment are shown in
Fig.~\ref{fig:rg-rules}. Like the standard definitions of rely
guarantee, our definition also relies on the \emph{stability}
condition, which requires pre and post conditions to hold despite the
interference from concurrent threads (captured by $R$). The stability
of pre and post conditions can be predicated on the assumption that
the interference preserves the trace invariant $\I$. Formally:
\begin{smathpar}
\begin{array}{lcl}
  \underI{\stable(R,P)} & \defeq & \forall \E, \E'.\, 
  \I(\E) \conj P(\E) \conj R(\E,\E') \\
  &   & \hspace*{1in}\conj \I(\E') \Rightarrow P(\E')\\
\end{array}
\end{smathpar}
However, the assumption that the interference preserves $\I$ (or,
equivalently, $\I$ withstands inteferference) needs to
be justified separately. We call this the stability requirement on
$\I$:
\begin{smathpar}
\begin{array}{lcl}
\stable(R,\I)& \Leftrightarrow & \forall \E, \E'.\, 
  \I(\E) \conj R(\E,\E')\Rightarrow \I(\E')\\
\end{array}
\end{smathpar}

The rule \rulelabel{RG-Var} defines the rely-guarantee judgment for
shared variable reads inside a transaction $T_i$. It requires $\I$ to
be stable, and pre and post conditions to be stable relative to $\I$.
The quantified premise effectively requires a proof that if the
abstract machine of Fig.~\ref{fig:txnimp} takes a step starting from
an execution $\E$ that satisfies the precondition $P$, then the
resultant execution $\E'$ satisfies the postcondition $Q$, and that
the guarantee $G$ faithfully captures the transition from $\E$ to
$\E'$. It is informative to compare this premise with the premise of
the \rulelabel{E-Aux} reduction rule of Fig.~\ref{fig:txnimp}. Similar
premises also appear in \rulelabel{RG-Asgn} and \rulelabel{RG-Txn}
rules, which define rely-guarantee judgments for assignments and
transactions, respectively. \rulelabel{RG-Var} however also requires
the return value ($\interp{S}(X)$) of the read to satisfy the
assertion $C$ meant for the value. \rulelabel{RG-Arith} defines the
RG judgment for an arithmetic expression $e_1\pm e_2$ in terms of the
corresponding judgments for the constitutent expressions $e_1$ and
$e_2$. Note that the pre-condition of $e_2$ ($Q'$) is also the
postcondition of $e_1$, reflecting the order of evaluation established
in Fig.~\ref{fig:txnimp}. The quantified premise requires any value
resulting from evaluating $e_1 \pm e_2$ to satisfy the assertion $C$,
provided that $e_1$ and $e_2$ always evaluate to values that satisfy
$C_1$ and $C_2$, respectively.

RG judgment for the assignment $X:=e$ makes use of the corresponding
judgment for the RHS expression $e$. The quantified premise asserts
that evaluating the assigment after evaluating $e$ to a value $v$ and
an execution $\E$, should result in an execution $\E'$ that satisfies
the postcondition $Q$, while being related to $\E$ via the guarantee
$G$. RG judgment for the \C{txn} lexical block is similar.  It uses
the judgment for the transaction-bound command $c$ (i.e.,
$\txnbox{c}_i$) to obtain an invariant $Q'$ for the execution $\E$
before the commit, and verifies that committing the transaction under
$\E$ results in an execution $\E'$ that satisfies transaction's
postcondition $Q$. As usual, $\E$ and $\E'$ need to be related by $G$.
The rules for sequential composition of transaction-bound comands
(\rulelabel{RG-Seq}) and parallel composition of top-level commands
(\rulelabel{RG-Par}) are straightforward, and more-or-less same as the
corresponding rules in classical rely-guarantee.

The three \rulelabel{RG-Conseq} rules define ways to strengthen or
weaken relations and assertions associated with RG judgments of
transaction-bound expressions, commands and top-level commands,
respectively. In particular, they allow the precondition $P$ and the
rely relation $R$ to be strengthened\footnote{A stronger relation is a
smaller set than a weaker relation, hence a stronger $R$ admits lesser
interference than a weaker $R$}, and the postcondition $Q$ (also, $C$
in the case of expressions) and the guarantee relation $G$ to be
weakened. This kind of strengthening/weakening is also admitted in the
classical RG framework. The most notable aspect of
\rulelabel{RG-Conseq} rules however is that they allow the trace
invariant $\I$ to be strengthened. Considering that $\I$ captures
consistency and isolation properties, this means that a program proved
correct under weaker consistency and isolation levels is also correct
under stronger levels.  Parametricity over the trace invariant $\I$,
combined with the ability to strengthen $\I$ as needed, allows our
proof system to admit a highly flexible proof strategy to prove
programs correct on various stores. Unlike the existing approaches,
which require programs to be proved correct with respect to a
pre-defined store (\eg~A CC store in~\cite{gotsmanpopl16}), our
reasoning framework allows programmers to \emph{define} a store of
their choice (by defining $\I$ appropriately) and then prove programs
correct with respect to this store. Thus, programmers can prove their
programs correct with respect to an imaginary store that provides only
the required consistency and isolation guarantees. Soundness of
strengthening $\I$ guarantees that the program can now be executed
safely on any real store (\eg~a relational database, or a NoSQL store)
that offers same or stronger consistency and isolation guarantees as
the imaginary store. The following subsection proves the soundness of
our proof system.

\subsection{Semantics and Soundness}

We now formalize the semantics of the rely-guarantee judgment of
Fig.~\ref{fig:rg-rules} through a series of definitions.

\begin{definition}[Step-indexed reflexive transitive closure]
Forall $A$, $R: A \rightarrow A \rightarrow \mathbb{P}$, and $n :
\mathbb{N}$, step-indexed reflexive transitive closure $R^n$ of $R$ is
the smallest relation satisfying the following
properties:
\begin{itemize}
\item $\forall (x:A).\, R^0 (x,x)$
\item $\forall (n:\mathbb{N})(x,y,z : A).\, R(x,y) \conj R^n(y,z) \Rightarrow
R^{n+1}(x,z)$
\end{itemize}
\end{definition}

\begin{definition}[Interleaved step and multi-step relations]
Interleaved step relation interleaves the thread-local reduction with
the interference from concurrent threads captured as the rely relation
($R$).  It is defined as following:
\begin{smathpar}
\begin{array}{lcl}
\I \vdash (c,\E) \rstepsto (c',\E') & \defeq & \I \vdash 
  (c,\E) \stepsto (c',\E') \\
  &   & \disj (c' = c \conj R(\E, \E') \conj \I(\E'))\\
\end{array}
\end{smathpar}
The interleaved step relation for transaction bound expressions
($\txnbox{e}_i$) and commands ($\txnbox{c}_i$) is defined similarly.
Interleaved multi-step relation ($\stepssto{n}$) is the step-indexed
reflexive transitive closure of the interleaved step relation.
\end{definition}

\begin{definition}[Semantics of the RG judgment]
Semantics of the RG sextuple $\rg{\mathbb{I},P,R}{c}{G,Q}$ is defined
in terms of the interleaved step relation as following:
\begin{smathpar}
\begin{array}{l}
\hspace*{-0.3in}
\rg{\mathbb{I},P,R}{c}{G,Q} \;\defeq\; \forall \E.\, P(\E)
  \wedge \mathbb{I}(\E) \\
\hspace*{0.4in}\Rightarrow (\forall n,\E'.\; \I \vdash (c,\E) 
    \rstepssto{n} (\cskip,\E') \Rightarrow Q(\E')) \\
\hspace*{0.5in}\conj \texttt{step-guaranteed}(\I,R,G,c,\E)\\
\end{array}
\end{smathpar}
The first conjunct in the consequent is called the \emph{Hoare
consequent} since it ascribes Hoare triple semantics to RG sextuple.
The second conjunct, called the \emph{guarantee consequent}, uses the
$\texttt{step-guaranteed}$ predicate defined below:
\begin{smathpar}
\begin{array}{l}
\texttt{step-guaranteed}(\I,R,G,c,\E) \;\defeq\; \forall n,\E',c'',\E''.\\
\hspace*{0.2in}\I \vdash (c,\E) \rstepssto{n} (c',\E') \conj \I \vdash (c',\E') \stepsto
  (c'',\E'') \Rightarrow G(\E',\E'')\\
\end{array}
\end{smathpar}
The guarantee consequent requires $G$ to capture the trace effect of
every small-step reduction of $c$, where reduction can be interleaved
by the interference ($R$) from concurrent threads. Semantics of the RG
sextuple for transaction-bound commands ($\txnbox{c}_i$) is defined
similarly. Expressions, unlike commands, evaluate to a value $v$, and
the semantics of their RG septuple
($\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$) differs slightly in that its
Hoare consequent requires the value $v$ to satisfy the assertion $C$. 
\end{definition}

It is worthwhile to note that the semantics of all RG judgments,
including the judgments for transaction-bound terms,  make similar
demands of the guarantee relation. Considering that transactions are
atomic (though not isolated), at a first glance, it might seem odd
that a transaction's guarantee is required to make explicit every step
of its reduction. This requirement is however justifed because, in
reality, a transaction's atomicity is predicated on the isolation
settings of the observer. A \iso{Read Uncommitted} transaction, for
example, can observe the internal state of a transaction, regardless
of the latter's preferences\footnote{Thus, there is no such thing as
atomicity; it's all about isolation.}. In the interest of modular
verification, the transaction must therefore make its internal state
available via its guarantee relation.

Soundness of rely-guarantee judgments can now be stated
thus\footnote{Supplemntary contains a rigorous proof of the soundness
theorem, and all relevant lemmas.}:
\begin{theorem}[Soundness] 
Rely-guarantee judgments defined by the rules in
Fig.~\ref{fig:rg-rules} are sound with respect to their semantics
defined above. 
\end{theorem}
\noindent In particular, if $\rg{\I,P,R}{c}{G,Q}$ can be derived using
the rules of Fig.~\ref{fig:rg-rules}, then (a). every interleaved
multi-step reduction of $c$ starting from a trace that satisfies $P$
and $\I$, results in a trace that satisfies $Q$, and (b). the effect
that every small-step reduction of $c$ along the way has on the trace
is contained in $G$. Soundness of RG judgment for transaction-bound
commands ($\txnbox{c}_i$) is stated similarly.  For expressions,
soundness of the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also
proves that $e$ is always evaluated to a value that satisfies $C$.

\subsection{Soundness under Eventually Consistent Replication}

\GK{Here, I give intuitive explanation of soundness in the context of
EC replication. This answers Gustavo's question of why Gotsman-like
special case reasoning about replicas is not needed in our setting.}

\subsection{Examples}

\GK{Here, I plan to include the full example that we dropped from
\S\ref{sec:motivation}.}
